---
title: "Loops in R"
subtitle: "<br>*Introduction to Bioinformatics - Module 2*"
author: "Nina Norgren"
format: 
    revealjs:
        theme: [serif, ../../../custom.scss]
        css: ../../module2.css
embed-resources: true
---

## What are loops and why do we need them?

<br>
```{r, echo=T}
namelist <- c("Steve", "Carol", 
              "Jonas", "Sarah")
```

## What are loops and why do we need them?
<br>

Print the names in the list one by one:

```{r, echo=T}
#| results: hold
print(namelist[1])
print(namelist[2])
print(namelist[3])
print(namelist[4])
```

## What are loops and why do we need them?
<br>

Print the names in the list one by one:

```{r, echo=T}
#| results: hold
print(namelist[1])
print(namelist[2])
print(namelist[3])
print(namelist[4])
```

Not very efficient...


## For loops

<br>
<br>
A `for` loop is a **control flow** statement that performs a **fixed operation** over a **known amount** of steps.

## Syntax of a for loop
<br>

```
for (<value> in <sequence>) {
    <action on value>
}
```

## Syntax of a for loop
<br>

```{r, echo=T}
namelist <- c("Steve", "Carol", 
              "Jonas", "Sarah")

for (name in namelist) {
    print(name)
}
```

## Syntax of a for loop
<br>

```{r, echo=T}
namelist <- c("Steve", "Carol", 
              "Jonas", "Sarah",
              "José")

for (name in namelist) {
    print(name)
}
```

## What can I loop over?
<br>

*Any sequence:*  

* Vectors  
* Lists  
* Matrices  
* Dataframes  
* Arrays  

## Examples - counters
<br>

```{r, echo=T}
#| results: hold
counter <- 0
for (name in namelist) {
    print(paste('Hello', name))
    counter <- counter + 1
}
print(paste('There are', counter, 
            'names in the list'))
```

## Examples - index positions
```{r, echo=T}
#| results: hold
for (i in 1:length(namelist)) {
    print(paste(namelist[i], 
                "has position", i))
}
```

## Examples - data frames

```{r, echo=F}
df <- data.frame(c(1:3),
                 c('a','b','c'),
                 c(TRUE, FALSE, TRUE))
colnames(df) <- c('NO.', 'char', 'isBlue')
df
```
<br>
Loop over columns (default):
```{r, echo=T}
for (column in df) {
    print(column)
}
```

## Examples - data frames

```{r, echo=F}
df <- data.frame(c(1:3),
                 c('a','b','c'),
                 c(TRUE, FALSE, TRUE))
colnames(df) <- c('NO.', 'char', 'isBlue')
df
```
<br>
Loop over rows:
```{r, echo=T}
for (i in 1:nrow(df)) {
    print(df[i,])
}
```


## While loops

<br>
<br>
A `while` loop is a **control flow** statement that allows code to be executed **repeatedly** based on a given **Boolean condition**.

## Syntax of a while loop
<br>

```
while (<condition is true>) {
    <action on value>
}
```

## Syntax of a while loop
<br>

```{r, echo=T}
#| results: hold
x <- 0
while (x < 5) {
   print(x)
   x <- x + 1
}
```

## Syntax of a while loop
<br>

Beware of infinite loops!
```{r, echo=T, eval=F}
#| results: hold
x <- 0
while (x < 5) {
   print(x)
}
```

## When to use what
<br>

* For loops better for simple iterations over vectors and other iterable objects
* While loops are more flexible and can iterate an unspecified number of times

## lapply
<br>

* R has a built in vectorized "for loop", called `lapply`
* `lapply` applies a function over a list and returns the result

```{r, echo=T}
#| results: hold
lapply(namelist, print)
```

## Summary
<br>

* For loops are used to iterate over a fixed number of steps in a sequence
* While loops will continue to iterate while a condition holds true