---
title: "if/else statements and functions"
subtitle: "<br>*Introduction to Bioinformatics - Module 2*"
author: "Nina Norgren"
format: 
    revealjs:
        theme: [serif, ../../../custom.scss]
        css: ../../module2.css
embed-resources: true
---

## What are if/else statements?
<br>

* Conditionals that only execute if a statement is true
* Allows us to execute different actions based on the condition

## Syntax of an if/else statement
<br>
```
if (<boolean is true>) {
    <action>
} else {
    <another action>
}
```

## Example
<br>
```{r, echo=T}
numbers <- c(1,3,5,7)

if (sum(numbers) > 5) {
    print("the sum is higher than 5")
} else {
    print("the sum is lower than 5")
}
```

## Example
<br>
```{r, echo=T}
numbers <- c(1,3,5,7)

if (sum(numbers) > 25) {
    print("the sum is higher than 5")
} else {
    print("the sum is lower than 5")
}
```

## Example - within loop
<br>
```{r, echo=F}
numbers <- matrix(1:9,
           nrow=3,
           byrow=T)
numbers
```

In what rows is the sum bigger than 15?

## Example - within loop
<br>
```{r, echo=F}
numbers <- matrix(1:9,
           nrow=3,
           byrow=T)
numbers
```

In what rows is the sum bigger than 15?

```{r, echo=T}
for (row in 1:nrow(numbers)) {
    if (sum(numbers[row,]) > 15) {
        print(numbers[row,])
    }
}
```



## What are functions?

<br>

Reusable blocks of code that performs a **specific task** and can be called upon whenever needed within a program

## What are functions?

<br>

Reusable blocks of code that performs a **specific task** and can be called upon whenever needed within a program

<br>

```{r, echo=T}
sd(c(1,2,3,4,5))
```

A built-in function to calculate standard deviation

## What are functions?

<br>

Reusable blocks of code that performs a **specific task** and can be called upon whenever needed within a program

<br>

```
function (x, na.rm = FALSE) {
  sqrt(var(if (is.vector(x) || is.factor(x)) x else as.double(x), 
           na.rm = na.rm))
}
```

## Syntax
<br>
```
my_func <- function(<input parameters>) {
    <action>
    return <return value>
} 
```
## Examples
<br>

```{r, echo=T}
second_item <- function(my_list) {
    second <- my_list[2]
    return (second)
}

result <- second_item(c(1,5,3,6))
print(result)
```

## Examples
<br> 

```{r, echo=T}
find_ten <- function(my_list) {
    count <- 0
    for (x in my_list) {
        if (x == 10) {
            count <- count + 1
        }
    }
    return (count)
}
result <- find_ten(c(1,4,10,5,10))
print(result)
```


## Why do we use functions
<br>

* Cleaner code
* Better defined tasks in code
* Re-usability
* Better structure


## Summary
<br>

* if/else statements allows us to execute different code based on a condition
* a function is a block of code that does one specific task, and can be re-used throughout the code 