---
title: "Variables"
author: "Nina Norgren"
format: 
    revealjs:
        theme: [serif, ../../../custom.scss]
        css: ../../module2.css
embed-resources: true
---

## R as a calculator

Let's start of easy:

```{r,echo=TRUE,size='Tiny'}
7 + 9
```

## R as a calculator

Let's start of easy:

```{r,echo=TRUE,size='Tiny'}
7 + 9
```
**Not very useful so far**


## Introducing: Variables

Creating a variable is simply assigning a name to some structure that stores data

```{r,echo=TRUE,size='Tiny'}
7 + 9
```

## Introducing: Variables

Creating a variable is simply assigning a name to some structure that stores data

```{r,echo=TRUE,size='Tiny'}
7 + 9
```
```{r,echo=TRUE,size='Tiny'}
a <- 7
b <- 9
a + b
```

---
## Variables cont.

We are not constrained to only numbers:

```{r,echo=TRUE}
text1 <- 'a'
text2 <- 'qwerty'
text1  
text2
```

## Is `<-` equivalent to `=`?

`val <- 3`, `val = 3` and `3 -> val` are three equivalent ways of assigning in R. 

But, it's best to use only `<-` to avoid possible confusion.


## Naming conventions

- How to write variable names?
- What is legal/valid/good style?

## Naming conventions

- How to write variable names?
- What is legal/valid/good style?

A valid name consists of:  

- letters
- numbers
- dot
- underline
- starts with a letter or the dot not followed by a number
- can not be a reserved keyword

## Examples

#### Allowed
my_data  
my.data  
.mydata  

## Examples

#### Allowed
my_data  
my.data  
.mydata  

<br>

#### Not allowed
.2data (starts with dot followed by number)  
2data (starts with a digit)  
_data (starts with an underscore)  
my data (has a blank in the middle)  
function (reserved keyword)  

## Reserved keywords

```
if, else, repeat, while, function, for, in, next, break, TRUE, FALSE, NULL, Inf, NaN, NA, NA_integer_, NA_real_, NA_complex_, NA_character_
```
<br>

- and you also **cannot** use: `c, q, t, C, D, I`
- and you **should not** use: `T, F`

## Good variable names

- Make them informative, e.g. `genotypes` instead of `fsjht45jkhsdf4`
- Use the same *naming convention* across your code:
  - camelNotation
  - dot.notation
  - dash_notation

- Do not `give_them_too_long_names`,

- In the dot notation avoid `my.variable.2`, use `my.variable2` instead