---
title: "Complex data structures: Data frames"
author: "Nina Norgren"
format: 
    revealjs:
        theme: [serif, ../../../custom.scss]
        css: ../../module2.css
embed-resources: true
---

## Complex data structures

dim | Homogenous | Heterogenous
----|------------|-----------------
1d  | vectors    | list
2d  | matrices   | data frame
nd  | arrays     | n/a  

## Data frames  

A **data frame** or a **data table** is similar to a matrix. In this structure elements of every column have the same type, but different columns can have different types. 

```{r, echo=T}
df <- data.frame(c(1:3),
                 c('a','b','c'),
                 c(TRUE, FALSE, TRUE))
df
```

## Data frames

Assign names to columns for readability

```{r, echo=T}
df <- data.frame(no.=c(1:3),
                 letter=c('a','b','c'),
                 isRed=c(TRUE, FALSE, TRUE))
df
```

## Data frames - accessing values

```{r, eval=TRUE, echo=FALSE}
df <- data.frame(no.=c(1:3),
                 letter=c('a','b','c'),
                 isRed=c(TRUE, FALSE, TRUE))
df
```

## Data frames - accessing values

```{r, eval=TRUE, echo=FALSE}
df <- data.frame(no.=c(1:3),
                 letter=c('a','b','c'),
                 isRed=c(TRUE, FALSE, TRUE))
df
```

<br>

First row:
```{r, echo=T}
df[1,]
```

## Data frames - accessing values

```{r, eval=TRUE, echo=FALSE}
df <- data.frame(no.=c(1:3),
                 letter=c('a','b','c'),
                 isRed=c(TRUE, FALSE, TRUE))
df
```

<br>

Second column:
```{r, echo=T}
df[,2] 
```

## Data frames - accessing values

```{r, eval=TRUE, echo=FALSE}
df <- data.frame(no.=c(1:3),
                 letter=c('a','b','c'),
                 isRed=c(TRUE, FALSE, TRUE))
df
```

<br>

Rows 2-3 from the isRed column:
```{r, echo=T}
df[2:3, 'isRed']
```

## Data frames - accessing values

```{r, eval=TRUE, echo=FALSE}
df <- data.frame(no.=c(1:3),
                 letter=c('a','b','c'),
                 isRed=c(TRUE, FALSE, TRUE))
df
```

<br>

The first 2 characters in letter:
```{r, echo=T}
df$letter[1:2]
```

## Data frames - merging

A very useful feature of R is merging two data frames on certain key using  **merge**:

```{r, echo=T}
df1 <- data.frame(no=c(1:3),
                 letter=c("a","b","c"))
df2 <- data.frame(no=c(1:3),
                 letter=c("A","B","C"))
merge(df1, df2, by='no')
```

## What else? stringsasfactors?