---
title: "Plotting with base graphics"
author: "Nina Norgren"
format: 
    revealjs:
        theme: [serif, ../../../custom.scss]
        css: ../../module2.css
embed-resources: true
---

## Graphical devices

The concept of a **graphical device** is crucial for understanding R graphics.  


A device can be a screen (default) or a file.

## Creating a plot 

Creating a plot entails:

* opening a graphical device (not necessary for plotting on screen),

* plotting to the graphical device

* closing the graphical device (very important!)


## The most commonly used graphical devices are:

* screen
* bitmap/raster devices: `png()`, `bmp()`, `tiff()`, `jpeg()`
* vector devices: `svg()`, `pdf()`


## Working with graphical devices

```{r, eval=FALSE, echo=TRUE}
png(filename = 'myplot.png', width = 320, 
    height = 320)
plot(x=c(1,2,7), y=c(2,3,5))
dev.off()
```

::: {.fragment}
What we did was, in sequence:


* open a graphical device, here `png()` with some parameters
:::

::: {.fragment}
* do the actual plotting to the device using `plot()` 
:::

::: {.fragment}
* close the graphical device using `dev.off()`.
:::

## 

> It is of paramount importance to remember to close graphical devices. Otherwise, our plots may end up in some random weird files or on screens.


## Scatter plot 

```{r, echo=TRUE}
x <- c(1, 2, 3, 4, 5)
y <- c(2, 4, 1, 6, 3)
plot(x, y, main = "Example",
     xlab = "X-axis",            
     ylab = "Y-axis",          
     pch = 16, col = "blue")
```

## Bar plot

```{r, echo=TRUE}
categories <- c("A", "B", "C", "D")
values <- c(10, 15, 7, 20)
barplot(values, 
        names.arg = categories, 
        main = "Bar Plot Example",
        xlab = "Category",
        ylab = "Value", col = "green")
```

## Box plot 

```{r, echo=TRUE}
group1 <- c(23, 28, 32, 35, 18)
group2 <- c(15, 20, 28, 30, 25)
group3 <- c(10, 15, 22, 25, 30)
data <- list(group1, group2, group3)
boxplot(data, 
        names = c("Group 1", "Group 2", "Group 3"),
        main = "Box Plot Example",
        xlab = "Groups",
        ylab = "Values",
        col = c("blue", "green", "pink"))
```

## Base graphics limitations

* Inconsistent grammar
* Mixing of aesthetics
* Hard to learn

::: {.fragment}
Use `ggplot2` instead!
:::